---
title: "Getting_started.Rmd"
output: html_document
date: "2022-12-18"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Intro

crosshap (R package) is an LD-based local haplotyping tool designed to visualize 
patterns of variation across loci and individuals in a genomic region of interest. 
In this tutorial we'll walk through an example analysis that will help to 
introduce and describe the main functions of crosshap so you'll be ready to jump
into exploring your own data.

The data we'll be analyzing is soybean data from the USDA core collection 
(published in Bayer et al. 2021, doi: 10.1002/tpg2.20109). This tutorial will
re-perform local haplotype analysis of the seed protein QTL, cqProt-003, with 
crosshap (see Marsh et al. 2022, doi: 10.1007/s00122-022-04045-8). 

``` {r installation, eval = FALSE}
#install.packages("devtools")
devtools::install_github("JacobIMarsh/crosshap")
```

``` {r message = FALSE, warning = FALSE, eval = TRUE}
library(crosshap)
```

## Preparing your inputs

For all inputs a wrapper is provided in crosshap to help read them in consistently. 
You can just provide the path to the files for your data, but for this tutorial 
we need to find the example data within the local cloned directory.

``` {r findpaths, eval = TRUE, include = T}
#Find path to example data
path_to_vcf <- system.file('extdata','imputed_173kb.vcf', package = 'crosshap')
path_to_LD <- system.file('extdata','plink.ld', package = 'crosshap')
path_to_pheno <- system.file('extdata','protein_phenotypes.txt', package = 'crosshap')
path_to_metadata <- system.file('extdata','domestication.txt', package = 'crosshap')
```

The VCF contains genomic variation (SNP) information of 985 accessions. Only 
sites within the 173kb trait associated region of interest are present in the 
VCF, along with header information labeling each individual. 

The data has been filtered and imputed; currently only biallelic variant sites
are supported by crosshap. We suggest you filter your data prior to analysis, 
though imputation is optional and often it will be preferable to keep it unimputed.
#Read into R using read_vcf()


```{r read_vcf, eval = T, include = T}
test_vcf <- read_vcf(path_to_vcf)
```

```{r print_vcf, eval = T, include = T}
head(test_vcf)
```


Next, the LD matrix provides pairwise linkage information between all SNP loci 
within the region of interest (173kb). For this example input a simple R^2 
matrix was calculated using PLINK (Purcell et al 2007): plink --r2 square --vcf
imputed_173kb.vcf.

The LD file must be a square correlation matrix using the same loci present in 
test_vcf, ordered by position, with or without position labels (vcf object must 
be provided if position labels are missing). Any LD metric that captures pairwise 
correlation on a scale from 0 (unlinked) to 1 (fully linked) may be used, including 
D', R^2, or R^2 corrected for kinship or population structure. 

```{r read_LD, eval = F}
#Read into R using read_LD()
test_LD <- read_LD(path_to_LD, vcf = test_vcf)
head(test_LD)
```

Providing phenotype scores for all, or at least some individuals in the VCF is 
useful for uncovering trait-associated genomic variation and haplotypes. Given 
this tutorial analyzes a seed protein QTL, seed protein % is the phenotype 
provided for this example. 

The phenotype file should be a comma or tab-delimited plain text file with 
individual names (same as VCF IDs) in the first column, and numerical phenotype 
scores in the second column. Note: this can be the same phenotype file provided 
for a GWAS.

```{r read_pheno, eval = F}
#Read into R using read_pheno()
test_pheno <- crosshap::read_pheno(path_to_pheno)
head(test_pheno)
```

Providing a metadata grouping label for distinct subpopulations in your 
dataset can aid in interpreting the haplotyping results. This example provides 
level of domestication for each individual, including undomesticated wild-type soy 
(wt), to domesticated landraces (lr), domesticated cultivars improved by advanced 
breeding (ocult) and further improved modern cultivars (mcult). 

The metadata file should be formatted the same as the phenotype file, though with
a categorical metadata grouping variable in the second column instead of numerical
phenotype score. This variable can be anything you want to check the spread of
across the haplotypes (e.g. geographic region), or that could be associated with 
phenotypic variation (e.g. year of collection).

``` r
#Read into R using read_read_metadata()
test_metadata <- read_metadata(path_to_metadata)
```

## Running local haplotyping

!! INTRO
Local haplotyping involves classifying individuals by shared haplotypes 
(combinations of alleles) across loci within a region of interest. In crosshap, 
haplotypes are not defined by individual loci but rather by allelic combinations 
of Marker Groups. Marker Groups are clusters of tightly linked SNPs which are 
typically inherited as a unit throughout the population. The clustering algorithm
used to capture Marker Groups is DBSCAN (Ester et al. 1996) which has two main 
parameters in crosshap: MGmin and epsilon.

You will need to pick a MGmin (minPts), which is the minimum number of linked SNPs 
in a cluster for it to be classified as a Marker Group instead of background noise. 
To give an approximate scale, a 300kb region with 3000 SNPs may be appropriately 
captured with a MGmin in the range of 20-40, depending on the desired granularity. 
In contrast, a 50kb region with 100 SNPs may work better with an MGmin of 5-15. In 
this example (2041 SNPs in 173kb region), a MGmin of 30 was chosen, though feel 
free to experiment to learn the effects of this parameter.

``` r
#Add minimum Marker Group SNP count
test_MGmin <- 30
```

Epsilon roughly translates to cluster density, when epsilon increases, more SNPs
will typically be included in more Marker Groups. At this stage it's worth providing a 
range of epsilons so you can pick the best one later. This dataset has a lot of SNPs 
(2041 loci), so the range of epsilons we will test will be quite low 
(0.2,0.4,0.6,0.8,1). Your dataset may require increasing these values 
(e.g. 1,1.5,2,2.5,3) if you want to capture more Marker Groups. Note: Epsilon values 
that are very high will begin decreasing the number of Marker Groups identified.

``` r
#Add list of epsilon values to test
test_epsilon <- c(0.2,0.4,0.6,0.8,1)
```
Now you've got all your inputs and parameters ready, it's time to run the haplotyping!

``` r
#Run the haplotyping at all provided epsilon values
run_haplotyping(vcf = test_vcf,
                LD = test_LD,
                pheno = test_pheno,
                metadata = test_metadata,
                epsilon = test_epsilon,
                MGmin = test_MGmin)
```

If everything went to plan, 5 Haplotype S3 objects should have been saved to your global
environment - one for each provided epsilon value. These files will be needed to build 
the crosshap visualizations. 

## Picking the optimal clustering resolution

Epsilon is a notoriously difficult parameter to optimize for DBSCAN, the tool used
to cluster SNPs into Marker Groups for haplotyping with crosshap. Thankfully, crosshap
has a clustree (Zappia et al 2018) wrapper that summarizes the differences between the 
haplotype objects created at the different epsilon values to help you pick the best one. 

The first clustering tree option provides a summary of changes in haplotype groups 
resolved at different epsilon values. 

``` r
#Provide phenotype data and parameters used to create haplotype objects
#Add type = 'hap' to ensure it summarizes haplotypes rather than Marker Groups
hap_clustree <- run_clustree(MGmin = test_MGmin,
                              pheno = test_pheno,
                              type = 'hap',
                              epsilon = test_epsilon)
hap_clustree
```

<details>
  <summary>**What are the circles and arrows on this plot?**</summary>
  
Each row of nodes represents the different haplotypes (A-Z) idenfied at a specific epsilon resolution (reported on right). The size of each node indicates the frequency of individuals that possess a given haplotype and the colour reflects the mean phenotype score for that population of individuals. The '0' node on the left represents the 
population of individuals that are not placed in a haplotype. The arrows between each row indicates that individuals in the upper haplotype node are present in the lower haplotype node, with the opacity reflecting the proportion of individuals. 

</details>

What you may notice is that the number of haplotypes identified changes from three 
populations at epsilon = 0.2 to six populations at epsilon = 0.6, though remains constant
with six haplotype populations as epsilon increases to 1. This is a good sign that the 
haplotypes identified at this epsilon range (0.6-1) are robust and effectively capture
a distinct genetic structure across the population in our region of interest. 

In addition, a range of colours are present for the rows of nodes corresponding to 
epsilons 0.6-1 which indicates there's phenotypic diversity between a large haplotype 
A which on average has a low phenotype score (seed protein %), and smaller haplotypes 
such as E and F which on average have higher phenotype scores.

Given these factors, lets pick the smallest 'stable' epsilon value, 0.6, to investigate 
further with the full crosshap visualization.

<details>
  <summary>**What if I want to visualize changes between marker groups?**</summary>
Some analyses will be more interested in the linked SNPs which may be causal for 
phenotypic trait variation, or may be markers for causal variants. In this case, 
picking a clustering resolution (epsilon) that is optimal for capturing informative
Marker Groups is the goal. The clustree wrapper provided by crosshap has a MG option 
for capturing changes between Marker Groups, rather than haplotypes, between results 
using different epsilon values. When "type = 'MG'", the nodes will instead represent 
Marker Groups, with size indicating number of SNPs, and colour reflecting the phenotypic 
association of a given Marker Groups. The '0' group represents background SNPs that 
aren't assigned to a Marker Group.

``` r
#Provide phenotype data and parameters used to create haplotype objects
#Add type = 'MG' to ensure it summarizes Marker Groups rather than haplotypes
MG_clustree <- run_clustree(MGmin = test_MGmin,
                              pheno = test_pheno,
                              type = 'MG',
                              epsilon = test_epsilon)
MG_clustree
```
</details>

### Visualizing local haplotypes with crosshap

Now that we have our ready as a haplotype object, and have picked the optimal epsilon 
resolution for our dataset, it's time to visualize our results in full so we can start 
uncovering the patterns of variation present in our region of interest!

The idea of the eponymous crosshap visualization is to integrate Marker Group results 
(cluster of SNPs) with features of haplotype populations (groups of genetically similar 
individuals), allowing you to intuitively connect genetic variants with the individuals 
that possess them. One of the reasons it's important to capture both in the same figure 
is because often trends of genetic variation captured by Marker Groups can be explained 
by population structure across the haplotypes - and vice versa. 

On the crosshap visualization, the central matrix displays the combinations of Marker 
Group alleles that define each haplotype, it provides a grid around which features of 
each Marker Group and haplotype combination is characterized Features of haplotypes and 
the individuals assigned to them are visualized vertically, with features in the top and 
bottom plots. Features of Marker Groups and the SNPs assigned to them are visualized 
horizontally, with features on the left and right plots. 

``` r
#Visualize haplotype object created by run_haplotyping() at the chosen epsilon (0.6)
Hap_viz <- crosshap_viz(Haplotypes_MGmin30_E0.6)
Hap_viz
#If it doesn't scale well in your plot viewer, try exporting it as an image with ggsave()
```

## What's shown in the crosshap visualization

Having a look at the haplotype visualization, there's a lot going on at first glance, 
so let's break it down piece by piece. In the central matrix we can see there are 9 
Marker Groups (MG1-MG9) defining 6 haplotypes (A-F). 

At the top, above the central matrix is the frequency of individuals with each haplotype 
combination, with a breakdown by our provided metadata variable (level of domestication).

At the bottom, below the central matrix is plotted the phenotype scores (seed protein %) 
for each individual (each dot is an individual), grouped by haplotype and coloured by 
metadata variable, like the top bar plot.

To the left of the central matrix is a summary of allele frequency information for each 
of the Marker Groups (MG1-MG9). This indicates the frequency of minor allele, missing 
alleles, heterozygotes, and reference alleles averaged across all the SNPs within each 
Marker Group.

To the right of the central matrix is plotted the phenotypic association for each SNP 
(each dot is a SNP), grouped by Marker Group. The 'pheno association' is effectively the 
difference in phenotype scores between individuals with the alternate compared to 
reference alleles for a given SNP.

The table in the bottom right summarizes the haplotype information and the table in the 
top right summarizes the Marker Group information. All guides are shown in the bottom 
left.

>>Are there other options for these plots?
The left plot providing allele frequency information can be swapped out by providing the 
argument "plot_left = 'pos'" to instead report the chromosomal position of each SNP in 
the region of interest. Similarly. the right plot providing phenotypic association 
results for SNPs can be swapped using "plot_right = 'cluster'" to instead build a 
boxplot which indicates the level of internal linkage within each Marker Group.

## Interpreting the crosshap visualization

Alright, now that we know how to read the crosshap visualization, lets actually use it 
to uncover some biological findings to put in that paper/report/grant application. Give 
it a crack yourself, and if you get stuck, have a look at the example findings below.

>>Level of domestication across haplotypes
Looking at the top bar plot and the summary table in the bottom right, haplotype A 
almost exclusively contains domesticated individuals including the vast majority of 
old and modern improved cultivars. Haplotypes B, D and E primarily contain wild-type 
individuals and haplotypes C and F appear landrace-specific.

>>Phenotypic variability across haplotypes
Looking at the bottom phenotype plot, individuals in haplotypes B-F (which share the 
alternate allele for MG1) typically have higher seed protein scores than individuals in 
haplotype A. This trend is not only found in the wild-type specific haplotypes (B,D,E) 
but also the landrace-specific haplotypes C and F indicating there are some domesticated 
subpopulations that may share a causal high-protein mutation in the region.

>>Linkage patterns between Marker Groups
By looking at the central MG/hap matrix, we can see that all other Marker Groups have 
'nested' linkage to MG1, meaning the alternatealleles for MG2-MG9 SNPs tend to only be 
present when the alternate allele for MG1 is present. This may indicate a bottleneck, in 
this case due to artificial selection for a variant tightly linked to MG1, leading to 
dramatically reduced genetic variation in the major domesticated haplotype A.

>>Phenotypic association of Marker Groups
Looking at the pheno association plot on the right, several Marker Groups appear to be 
trait-associated, particularly MG1 and MG8. While MG8 is associated with the highest 
phenotypic effect (4.39), it has a low minor allele frequency (0.03; see left allele 
count plot). In contrast, MG1 has a substantially higher minor allele frequency (0.21), 
which means the phenotypic association for MG1 has more statistical power. 

>>Candidates for further analysis and validation
MG1 is the strongest candidate Marker Groups for further analysis, it may be worth 
investigating the functional annotations of these markers to identify a candidate causal 
variant, or testing whether indels/structural variation is linked to MG1. MG8 is also a 
candidate for futher analysis, though the alternate allele for MG8 is only found in 
18 landrace individuals with very high seed protein scores. As a result, it may be  
fruitful to investigate the subpopulation of individuals in haplotype F: are they from 
the same geographic region? do they share any other unique biological features? These 
rare landrace individuals in haplotype F may be interesting germplasm for introgression 
studies.

Congratulations, you did your first crosshap analysis! Now you're ready and equipped to 
give it a go with your data :)

## Exporting raw results and individual plots

You can have a closer look at the saved results by looking at the elements within the 
haplotype object. This is ideal for exporting information about the individuals that 
possess each haplotype, and the SNPs within each Marker Group for further analysis.

``` r
#Print first lines of haplotype object buckets
#The Indfile reports the haplotype assigned to each individual
Haplotypes_MGmin30_E0.6$Indfile

#The Hapfile reports the identified haplotypes, and their Marker Group combinations
Haplotypes_MGmin30_E0.6$Hapfile

#The Varfile reports information for each SNP, including Marker Group assignments
Haplotypes_MGmin30_E0.6$Varfile
```

If you're interested in a specific plot from the crosshap visualization, though don't 
need the full integrated figure, you can build the plots one-by one and stitch them 
together as you like using the patchwork package in R (Pederson 2022).

Below we'll only include the top and bottom plots that are relevant that characterize 
our haplotype populations.

``` r
#Install patchwork if you don't have it using install.packages("patchwork")
library(patchwork)

#First lets build the top plot, keeping the guides with "hide_labels = F"
top <- build_top_metaplot(Haplotypes_MGmin30_E0.6, hide_labels = F)

#Next lets build the bottom plot, hiding the guides with "hide_labels = T"
bot <- build_bot_halfeyeplot(Haplotypes_MGmin30_E0.6, hide_labels = T)

#Now we can stitch them together with patchwork!
hap_plots <- patchwork::wrap_plots(top, bot) + patchwork::plot_layout(ncol = 1)
```
